---
title: "Test mixture model on the globe"
author: "Z Sha"
date: "26 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

## 0 Introduction

In this document, we test the mixture model on the globe. We use the land coastline as the boudary for the mixture Gaussian models. We assume the same Gaussian process on the land and another on the ocean.


## 1 Define Polygons and Simulate data
First we read in a low resolution coastline to separate the lands and oceans and then define the true process by a GMRF defined on the triangle mesh of a SPDE model.

We need the following packages and extra functions from the sourced file.
```{r init, include=TRUE, message = FALSE, warning=FALSE}
## load libraries and source codes
library(INLA); library(rgdal); library(maptools); library(GEOmap); library(rgl)
source("C:/ZSwork/glbm/Experiment1b/mixture_model/BarrierModel/functions-barriers-dt-models-march2017.R")
set.seed(18)
```

Load the coastline and define polygons.
```{r polygons, include=TRUE}
lands <- readOGR(dsn = "C:/ZSwork/glbm/Experiment2/ne_110m_land", layer = "ne_110m_land")
## Remove tiny islands
landsareas <- sapply(lands@polygons, function(x) slot(x, "area"))
lands2 <- lands[landsareas > 10,]

## Define the global polygons
globe_p <- Polygon(coords = cbind(c(-180, -180, 180, 180, -180), c(-90, 90, 90, -90, -90)))

## Define the ocean polygons by adding holes in the globe
land_holes <- lapply(lands2@polygons, function(x) x@Polygons[[1]])
land_holes <- lapply(land_holes, function(x) {x@hole <- TRUE;return(x)})
land_holes[[length(land_holes) + 1]]  <- globe_p
Ocean <- Polygons(land_holes, ID = "a")
Ocean <- SpatialPolygons(list(Ocean), proj4string =CRS("+proj=longlat"))
```

Generate the point locations for building up the mesh. We want the mesh to be dense over the oceans and sparse on the lands.
```{r points1, include = TRUE}
## First create uniformly distributed points on the sphere -- Fibonacci Points
#loc_sphere <- spsample(Ocean, n = 1000, type = "Fibonacci")
## Could use spsample but there is a bug of coordinate translation so use my own Fibonacci code

#### Generate Fibonacci points on the sphere 
fiboSphere2 <- function(N = 1000L, L0 = FALSE) {
  ## Reference (note that points generated from 2D are slightly different from 3D)
  ## Measurement of Areas on a Sphere Using Fibonacci and Latitude–Longitude Lattices (2010)
  phi <- (sqrt(5) + 1) / 2 - 1 # golden ratio
  ga <- phi * 2 * pi           # golden angle
  
  i <- seq(-N, N)
  P <- 2 * N + 1
  lat <- asin(2*i / P) * 180 / pi
  if(L0){
  lon <- ((2 * pi * i / phi) %% pi) * 360 / pi
  }else{
    lon <- ((2 * pi * i / phi) %% pi) * 360 / pi - 180
    }
  cbind(lon = lon, lat = lat)
}

fibo_points <- fiboSphere2(N = 6000)
## Remove points on the land
pinocean <- unlist(over(Ocean, SpatialPoints(coords = fibo_points, proj4string = CRS("+proj=longlat")), returnList=T))
fibo_inOceans <- fibo_points[pinocean,]
plot(Ocean)
points(fibo_inOceans, pch = ".")
```

Now generate sparse points on the land using the same trick.
```{r points2, include = TRUE}
fibo_points <- fiboSphere2(N = 500)

## Find points on the land
pinocean <- unlist(over(Ocean, SpatialPoints(coords = fibo_points, proj4string = CRS("+proj=longlat")), returnList=T))
fibo_inLands <- fibo_points[-pinocean,]
plot(Ocean)
points(fibo_inLands, pch = "*", col = 2)
points(fibo_inOceans, pch = ".")
```

Finally assmeble these points and define the coastlines to be interior lines.
```{r points3, include = TRUE}
## Combine the points
mesh_points0 <- rbind(fibo_inOceans, fibo_inLands)
## convert the longlat coordinates to xyz
mesh_points0_xyz <- do.call(cbind, Lll2xyz(lat = mesh_points0[,2], lon = mesh_points0[,1]))

## define the interior lies as the coastlines
coast_seg <- inla.sp2segment(lands2)
coast_xyz <- do.call(cbind, Lll2xyz(coast_seg$loc[,2], coast_seg$loc[,1])) # project longlat back to cartesian xyz
coast_seg$loc <- coast_xyz
## Now generate the initial mesh and plot
mesh0 <- inla.mesh.2d(loc = mesh_points0_xyz, cutoff = 0.02, max.edge = 0.5)

plot3d(coast_seg$loc, pch = "-", col = 2)
plot(mesh0, rgl = T, add = TRUE)

## Use these points to iterate and add in the coastline as interior lines
mesh <- inla.mesh.2d(loc = mesh0$loc, interior= coast_seg , cutoff = 0.02, max.edge = 0.5)
plot(mesh)
```

Plot the mesh and see check which triangles are inside the polygons
```{r mesh2, include = TRUE}
mesh = dt.mesh.addon.posTri(mesh, globe = TRUE)
# - Add on mesh$posTri
# - - contains the positions of the triangles

## checking which mesh triangles are inside the land
## First convert xyz to lonlat
Tlonlat <- Lxyz2ll(list(x = mesh$posTri[,1], y = mesh$posTri[,2], z = mesh$posTri[,3]))
mesh$Trill <- cbind(lon = Tlonlat$lon, lat =Tlonlat$lat)
TinOcean <- unlist(over(Ocean, SpatialPoints(coords=mesh$Trill, proj4string = CRS("+proj=longlat")), returnList=T))
TAll <- 1:mesh$t
TinLand <- TAll[-TinOcean]
Omega = dt.Omega(list(TinLand, 1:mesh$t), mesh)
Omega.SP = dt.polygon.omega(mesh, Omega, globe = TRUE)

plot(mesh, main ="Mesh and Omega", asp = 1)
plot(Omega.SP[[1]],  col='grey')
plot(Omega.SP[[2]],  col='lightblue')
plot(mesh, add=T)

```

Now use this mesh to build the model and simulate data.
```{r simData, include = TRUE}
## Create the precsion matrix function
Q.function = dt.create.Q(mesh, Omega)
ranges = c(10, 0.17)
# - the first range is for the barrier area
# - - it is not sensitive to the exact value here, 
# - the second range is for the normal area
Q = Q.function(theta = c(log(1), log(ranges)))
# - the precision matrix for fixed ranges

## Simulate the field
u = inla.qsample(n=1, Q=Q, seed = 3)
u = u[ ,1]

## A wrapper for plotting the result
local.plot.field = function(field, ...){
  xlim = c(0, 1); ylim = xlim;
  proj = inla.mesh.projector(mesh, xlim = xlim, 
                             ylim = ylim, dims=c(300, 300))
  field.proj = inla.mesh.project(proj, field)
  image.plot(list(x = proj$x, y=proj$y, z = field.proj), 
             xlim = xlim, ylim = ylim, ...)  
}

## Plot the simulated field
local.plot.field(u, main="The true (simulated) spatial field", asp = 1, zlim = c(-6, 2.5))
```

