---
title: "Test mixture model on the globe"
author: "Z Sha"
date: "26 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

## 0 Introduction

In this document, we test the mixture model on the globe. We use the land coastline as the boudary for the mixture Gaussian models. We assume the same Gaussian process on the land and another on the ocean.


## 1 Define Polygons and Simulate data
First we read in a low resolution coastline to separate the lands and oceans and then define the true process by a GMRF defined on the triangle mesh of a SPDE model.

We need the following packages and extra functions from the sourced file.
```{r init, include=TRUE, message = FALSE, warning=FALSE}
## load libraries and source codes
library(INLA); library(rgdal); library(maptools); library(GEOmap); library(rgl)
source("c:/ZSwork/glbm/Experiment1b//mixture_model/BarrierModel/functions-barriers-dt-models-march2017.R")
set.seed(18)
```

Load the coastline and define polygons.
```{r polygons, include=TRUE}
lands <- readOGR(dsn = "c:/ZSwork/glbm/Experiment2/ne_110m_land", layer = "ne_110m_land")
## Remove tiny islands
landsareas <- sapply(lands@polygons, function(x) slot(x, "area"))
lands2 <- lands[landsareas > 10,]

## Define the global polygons
globe_p <- Polygon(coords = cbind(c(-180, -180, 180, 180, -180), c(-90, 90, 90, -90, -90)))

## Define the ocean polygons by adding holes in the globe
land_holes <- lapply(lands2@polygons, function(x) x@Polygons[[1]])
land_holes <- lapply(land_holes, function(x) {x@hole <- TRUE;return(x)})
land_holes[[length(land_holes) + 1]]  <- globe_p
Ocean <- Polygons(land_holes, ID = "a")
Ocean <- SpatialPolygons(list(Ocean), proj4string =CRS("+proj=longlat"))
```

Generate the point locations for building up the mesh. We want the mesh to be dense over the oceans and sparse on the lands.
```{r points1, include = TRUE}
## First create uniformly distributed points on the sphere -- Fibonacci Points
#loc_sphere <- spsample(Ocean, n = 1000, type = "Fibonacci")
## Could use spsample but there is a bug of coordinate translation so use my own Fibonacci code

#### Generate Fibonacci points on the sphere
fiboSphere2 <- function(N = 1000L, L0 = FALSE) {
  ## Reference (note that points generated from 2D are slightly different from 3D)
  ## Measurement of Areas on a Sphere Using Fibonacci and Latitude–Longitude Lattices (2010)
  phi <- (sqrt(5) + 1) / 2 # golden ratio
  ga <- phi * 2 * pi           # golden angle

  i <- seq(-N, N)
  P <- 2 * N + 1
  lat <- asin(2*i / P) * 180 / pi
  if(L0){
  lon <- ((2 * pi * i / phi) %% pi) * 360 / pi
  }else{
    lon <- ((2 * pi * i / phi) %% pi) * 360 / pi - 180
    }
  cbind(lon = lon, lat = lat)
}

fibo_points <- fiboSphere2(N = 4000)
## Remove points on the land
pinocean <- unlist(over(Ocean, SpatialPoints(coords = fibo_points, proj4string = CRS("+proj=longlat")), returnList=T))
fibo_inOceans <- fibo_points[pinocean,]
plot(Ocean)
points(fibo_inOceans, pch = ".")
```

Now generate sparse points on the land using the same trick.
```{r points2, include = TRUE}
fibo_points <- fiboSphere2(N = 500)

## Find points on the land
pinocean <- unlist(over(Ocean, SpatialPoints(coords = fibo_points, proj4string = CRS("+proj=longlat")), returnList=T))
fibo_inLands <- fibo_points[-pinocean,]
plot(Ocean)
points(fibo_inLands, pch = "*", col = 2)
points(fibo_inOceans, pch = ".")
```

Finally assmeble these points and define the coastlines to be interior lines.
```{r points3, include = TRUE}
## Combine the points
mesh_points0 <- rbind(fibo_inOceans, fibo_inLands)
## convert the longlat coordinates to xyz
mesh_points0_xyz <- do.call(cbind, Lll2xyz(lat = mesh_points0[,2], lon = mesh_points0[,1]))

## define the interior lies as the coastlines
coast_seg <- inla.sp2segment(lands2)
coast_xyz <- do.call(cbind, Lll2xyz(coast_seg$loc[,2], coast_seg$loc[,1])) # project longlat back to cartesian xyz
coast_seg$loc <- coast_xyz
## Now generate the initial mesh and plot
mesh0 <- inla.mesh.2d(loc = mesh_points0_xyz, cutoff = 0.02, max.edge = 0.5)

plot3d(coast_seg$loc, pch = "-", col = 2)
plot(mesh0, rgl = T, add = TRUE)

## Use these points to iterate and add in the coastline as interior lines
mesh <- inla.mesh.2d(loc = mesh0$loc, interior= coast_seg , cutoff = 0.02, max.edge = 0.5)
plot(mesh)
```

Plot the mesh and see check which triangles are inside the polygons
```{r mesh2, include = TRUE}
mesh = dt.mesh.addon.posTri(mesh, globe = TRUE)
# - Add on mesh$posTri
# - - contains the positions of the triangles

## checking which mesh triangles are inside the land
## First convert xyz to lonlat
Tlonlat <- Lxyz2ll(list(x = mesh$posTri[,1], y = mesh$posTri[,2], z = mesh$posTri[,3]))
mesh$Trill <- cbind(lon = Tlonlat$lon, lat =Tlonlat$lat)
TinOcean <- unlist(over(Ocean, SpatialPoints(coords=mesh$Trill, proj4string = CRS("+proj=longlat")), returnList=T))
TAll <- 1:mesh$t
TinLand <- TAll[-TinOcean]
Omega = dt.Omega(list(TinLand, 1:mesh$t), mesh)
#Omega.SP = dt.polygon.omega(mesh, Omega, globe = TRUE, round = 2)
## Plot the result in 3d
plot(mesh, t.sub = Omega[[2]], col = "lightblue", rgl = TRUE )
plot(mesh, t.sub = Omega[[1]], col = "yellow",  rgl = TRUE, add = TRUE)
writeWebGL(dir = "~/glbm/Experiment1b/mixture_model/", 
           filename= "~/glbm/Experiment1b/mixture_model/mixtureMesh.html",  # save the plot as an html
           width = 600, reuse = TRUE)

## Plot the result in 2d
plot(mesh, t.sub = Omega[[2]], rgl = T)
```

Now use this mesh to build the model and simulate data.
```{r simData, include = TRUE}
## Create the precsion matrix function
Q.function = dt.create.Q(mesh, Omega)
ranges = c(5, 0.1)
# - the first range is for the barrier area
# - - it is not sensitive to the exact value here,
# - the second range is for the normal area
Q = Q.function(theta = c(log(1), log(ranges)))
# - the precision matrix for fixed ranges

## Simulate the field
u = inla.qsample(n=1, Q=Q, seed = 2017)
u = u[ ,1]

## A wrapper for plotting the result
local.plot.field = function(field, ...){
  proj = inla.mesh.projector(mesh,  projection = "longlat", dims = c(360,180), xlim = c(-180, 180), ylim = c(-90, 90))
  field.proj = inla.mesh.project(proj, field)
  image.plot(list(x = proj$x, y=proj$y, z = field.proj),
              ...)
}

## Plot the simulated field
local.plot.field(u, main="The true (simulated) spatial field", asp = 1)
plot(Ocean, add = TRUE)
```

Now sample observations. The real observations are usually all over the Ocean and the pseudo-observations are along the polygon boundaries.
```{r observations1, include = TRUE}
## sample points in the ocean
obs_ocean <- spsample(Ocean, n = 2000, type = "random")

## sample points on the polygon boundaries
landslines <- as(lands2, 'SpatialLines') 
obs_coast <- spsample(landslines, n = 200, type = "random") # not points more than specified

proj4string(obs_coast) <- proj4string(obs_ocean)
obs_all<- rbind(obs_ocean, obs_coast)
plot(obs_all)
obs_allxyz <- do.call(cbind, Lll2xyz(lat = obs_all@coords[,2], lon = obs_all@coords[,1]))

A1.data = inla.spde.make.A(mesh, obs_allxyz)
# - the projector matrix required for any spatial model
# - this matrix can transform the field-defined-on-the-mesh 
#   to the field-defined-on-the-data-locations

u.data1 = A1.data %*% u
# - project the field from the finite element  
#   representation to the data locations
df = data.frame(obs_allxyz) # - df is the dataframe used for modeling
names(df) = c('locx', 'locy', 'locz')
sigma.epsilon <- c(rep(0.2,2000), rep(0.05, length(obs_coast))) # - size of the iid noise in the Gaussian likelihood
df$y = drop(u.data1 + sigma.epsilon*rnorm(nrow(df)))
# - sample observations with gaussian noise

summary(df)
```

## 2 Inference -- stationary
```{r stationary, include=TRUE}
stk1 <- inla.stack(data=list(y=df$y), A=list(A1.data),
                  effects=list(s=1:mesh$n), 
                  remove.unused = FALSE, tag='est')

model.stat = inla.spde2.matern(mesh)
# - Set up the model component for the spatial SPDE model: 
#   Stationary Matern model
# - I assume you are somewhat familiar with this model

formula <- y ~ -1 + f(s, model=model.stat)
# - Remove the default intercept
# - - Having it in the stack instead improves the numerical 
#     accuracy of the INLA algorithm
# - Fixed effects + random effects

hyper <- list(prec = list(fixed = TRUE, initial = log(1)))
scales = 1/sigma.epsilon^2  
  
res.stationary <- inla(formula, data=inla.stack.data(stk1),
            control.predictor=list(A = inla.stack.A(stk1)),
            family = 'gaussian', scale = scales,
            control.family = list(hyper = hyper))

summary(res.stationary)
res_pars <- inla.spde2.result(res.stationary, "s", model.stat, do.transf=TRUE)

plot(res_pars$marginals.range.nominal[[1]], type = "l", main = "range")
plot(res_pars$marginals.variance.nominal[[1]], type = "l", main = "variance")

local.plot.field(res.stationary$summary.random$s$mean, asp =1, zlim = c(-25, 20),
          main="Spatial estimate with the stationary model -- mean")
plot(Ocean, add = TRUE)
plot(obs_all, pch = ".", add = TRUE)
local.plot.field(res.stationary$summary.random$s$sd, asp =1,  
          main="Spatial estimate with the stationary model -- uncertainty")
plot(Ocean, add = TRUE)

```

## 2 Inference -- mixture model
```{r barrier, include = TRUE}
Q.mixture = dt.create.Q(mesh, Omega, fixed.ranges = c(5, NA))
# - We fix the barrier range to a different value than we 
#   used for simulations
# - - Why? It does not matter, as long as it is 'small' 
#     the models are very
#     similar
# - - This shows that you do not need to know the 
#     true 'barrier range'!

log.prior = dt.create.prior.log.exp(prior.param = c(1,1))
# - The prior parameters are the lambdas in the exponential 
#   priors for standard deviation and inverse-range

mixture.model = dt.inla.model(Q = Q.mixture, log.prior=log.prior)

formula2 <- y ~ -1 + f(s, model=mixture.model)
# - The spatial model component is different from before
# - The rest of the model setup is the same! 
#   (as in the stationary case)
# - - e.g. the inla(...) call below is the same, 
#     only this formula is different

res.mixture = inla(formula2, data=inla.stack.data(stk1),
       control.predictor=list(A = inla.stack.A(stk1)),
       family = 'gaussian',
       control.family = list(hyper = hyper),
            scale = scales)

summary(res.mixture)
theta1 <- res.mixture$marginals.hyperpar$`Theta1 for s`
theta2 <- res.mixture$marginals.hyperpar$`Theta2 for s`

Vmar<- inla.tmarginal(exp, theta1)
Rmar <- inla.tmarginal(exp, theta2)

plot(Vmar, type = "l", main = "posterior varianace")
plot(Rmar, type = "l", main = "posterior range")

local.plot.field(res.mixture$summary.random$s$mean, asp = 1,zlim = c(-25, 20),
                 main="Spatial posterior for mixture model -- mean")
plot(Ocean, add = TRUE)
plot(obs_all, pch = ".", add = TRUE)

local.plot.field(res.mixture$summary.random$s$sd, asp = 1, 
                 main="Spatial posterior for mixture model -- uncertainty")
plot(Ocean, add = TRUE)
plot(obs_all, pch = ".", add = TRUE)

```