---
title: "Estimating GIA -- 3 Pseudo Polygons defining subset of sphere as modelling region"
author: "Z Sha"
date: "13 Nov 2017"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction

In this study, we apply the Bayesian data assimilation model on updatding the global GIA process by using GPS data. We use the soluton from the ICE6G-VM5 model as our prior mean for the GIA process. The GPS data are processed into yearly vertical bedrock movement from a selected global network. 

We have first used a [***global stationary model***](http://rpubs.com/zs_sz/gia01) and now we try to deal with the non-stationary of the process comes from experts' opinion that at some regions the GIA should be zero in theroy for sure. Three different approaches are used in modelling the non-stationarity:

(1) Use pseudo-polygon observations --- [***Polygon observation***](http://rpubs.com/zs_sz/gia02)

(2) Use pseudo-polygons to define the subset where GIA can be modelled as stationary process --- [***Subset model***](http://rpubs.com/zs_sz/gia03)

(3) Use psdeuo-polygons to define a mixture Gaussian process where the updating process has different spatial properties inside and outside of the pseudo-polygons --- [***Mixture model***](http://rpubs.com/zs_sz/gia04)

In the following, we demonstrate the first approach.

# Define the pseudo polygons and mesh

First we define the pseudo polygons with help of experts' opinion. More details about choosing and generating the polygons can be found [here](http://rpubs.com/zs_sz/PseudoObs02). The following chunk load the pseudo polygons using the ensemble mean of 8 GIA mode solutions for a given threshold value, say $0.3$.

```{r ini_poly, message = FALSE, warning=FALSE}
## load library and functions
library(sp); library(INLA); library(GEOmap); library(rgdal)
library(ggplot2); library(grid); library(gridExtra)
source("functions.R")
source("functions-barriers-dt-models-march2017.R")

## Load the pseudo polygon
#### 1 Load GIA prior
if(Sys.info()["sysname"] == "Windows"){
  zeroPolygon <- readOGR(dsn = "Z:/WP1-BHM/Experiment1b/shapefiles", layer = "zero03")
}else if(grep("Ubuntu",Sys.info()["version"]) == 1){
  zeroPolygon <- readOGR(dsn = "/home/zs16444/GMdata/shapefiles", layer = "zero03")
}else {
  zeroPolygon <- readOGR(dsn = "/./projects/GlobalMass/WP1-BHM/Experiment1b/shapefiles", layer = "zero03")
}
plot(zeroPolygon, col = "blue",  main = "zero regions -- threshold = 0.3")

## Remove polygons that are too small
zeroPolys <- zeroPolygon@polygons[[1]]@Polygons
polyareas <- sapply(zeroPolys, function(x) x@area)
polyholes <- sapply(zeroPolys, function(x) x@hole)

zeropolys2 <- zeroPolys[polyareas > 200 ] 
zeroPoly <- zeroPolygon
zeroPoly@polygons[[1]]@Polygons <- zeropolys2
plot(zeroPoly, col = "blue", main = "zero regions -- threshold = 0.3 (refined)")
```

Next we generate the mesh and separate the triangles in/out-side of the polygons. We spread fibonacci points on the sphere as initial locations of the triangle vertices. The number of the points controls the resolution. The goal is to produce a map with approximately one degree resolution, hence we also need a similar resolution for the triangulation. Not that the triangle edge be no larger than the process correlation length. A one degree grid is about $110 \times 110 km$ which corresponds to a correlation length about $110/6371 = 0.017$. The number of triangles should be $f = 180 * 360 = 64800$ corresponding to number of vertices of $v \approx (f+2)*2/5= 25921$.

```{r mesh, include=TRUE, cache=TRUE}
fibo_points <- fiboSphere(N = 3e4, LL = FALSE)
mesh <- inla.mesh.2d(loc = fibo_points, cutoff = 0.01, max.edge = 0.5)
summary(mesh)
```

Now separate the triangles by the pseudo-polygons.
```{r mesh2, include = TRUE, cache=TRUE}
mesh <- dt.mesh.addon.posTri(mesh = mesh, globe = TRUE) # Add on mesh$posTri: contains the positions of the triangles

## checking which mesh triangles are inside the land
## First convert xyz to lonlat
Tlonlat <- Lxyz2ll(list(x = mesh$posTri[,1], y = mesh$posTri[,2], z = mesh$posTri[,3]))
Tlonlat$lon <- ifelse(Tlonlat$lon >=0, Tlonlat$lon, Tlonlat$lon + 359)
mesh$Trill <- cbind(lon = Tlonlat$lon, lat =Tlonlat$lat)
TinPoly <- unlist(over(zeroPoly, SpatialPoints(coords=mesh$Trill), returnList=T))
TAll <- 1:mesh$t
ToutPoly <- TAll[-TinPoly]
Omega = dt.Omega(list(TinPoly, 1:mesh$t), mesh)

## Plot the result in 3d
plot(mesh, t.sub = Omega[[2]], col = "lightblue", rgl = TRUE )
plot(mesh, t.sub = Omega[[1]], col = "yellow",  rgl = TRUE, add = TRUE)
```

Now we need to build an mesh from the subset of triangles that can be used in INLA independently. Find the subset of the vertices, and change all indices, etc.
```{r mesh3, inlude = TRUE, cahce=TRUE}
mesh_outPoly <- mesh.sub(mesh, Omega, 2)
plot(mesh_outPoly, rgl = T)
plot(mesh_outPoly)
```


# Load Data

Load the GIA prior mean and GPS data and do the same prepreation as previous.
```{r load_data0, include=FALSE, eval = TRUE, cache = TRUE}
#### 1 Load GIA prior
if(Sys.info()["sysname"] == "Windows"){
  ice6g <- read.table("Z:/WP2-SolidEarth/BHMinputs/GIA/GIA_Pel-6-VM5.txt", header = T)
}else if(grep("Ubuntu",Sys.info()["version"]) == 1){
  ice6g <- read.table("~/GMdata/BHMinputs/GIA/GIA_Pel-6-VM5.txt", header = T)
}else{
  ice6g <- read.table("/./projects/GlobalMass/WP2-SolidEarth/BHMinputs/GIA/GIA_Pel-6-VM5.txt", header = T)
}

polycoords <- ice6g[,c(6:13, 6,7)] 
plist <- lapply(ice6g$ID, 
                function(x) Polygons(list(Polygon(cbind(lon = as.numeric(polycoords[x, c(1,3,5,7,9)]), 
                                                        lat = as.numeric(polycoords[x, c(2,4,6,8,10)])))), ID = x))
Plist <- SpatialPolygons(plist, proj4string = CRS("+proj=longlat"))

meshLL <- Lxyz2ll(list(x=mesh_outPoly$loc[,1], y = mesh_outPoly$loc[,2], z = mesh_outPoly$loc[,3]))
meshLL$lon <- ifelse(meshLL$lon >= -0.5, meshLL$lon,meshLL$lon + 360)
mesh_sp <- SpatialPoints(data.frame(lon = meshLL$lon, lat = meshLL$lat), proj4string = CRS("+proj=longlat")) 
mesh_idx <- over(mesh_sp, Plist)
GIA_prior <- ice6g$trend[mesh_idx]

#### 2 Load GPS data
if(Sys.info()["sysname"] == "Windows"){
  GPSV4b <- read.table("Z:/WP2-SolidEarth/BHMinputs/GPS/GPS_v04b.txt", header = T)
}else if(grep("Ubuntu",Sys.info()["version"]) == 1){
  GPSV4b <- read.table("~/GMdata/BHMinputs/GPS/GPS_v04b.txt", header = T)
}else{
  GPSV4b <- read.table("/./projects/GlobalMass/WP2-SolidEarth/BHMinputs/GPS/GPS_v04b.txt", header = T)
}
```

##  GPS data

Remove GPS data inside the pseudo-polygons.
```{r data, include = TRUE, cache=TRUE}
GPS_inPoly <- unlist(over(zeroPoly, SpatialPoints(coords = cbind(GPSV4b$lon, GPSV4b$lat)), returnList=T))
GPS_All <- 1:nrow(GPSV4b)
GPS_outPoly <- GPS_All[-GPS_inPoly]
plot(GPSV4b[GPS_outPoly,c("lon", "lat")], pch = "+")

GPS_data <- GPSV4b[GPS_outPoly,]
GPS_loc <- do.call(cbind, Lll2xyz(lat = GPS_data$lat, lon = GPS_data$lon))
GPS_sp <- SpatialPoints(data.frame(lon = ifelse(GPS_data$lon>359.5, GPS_data$lon - 360, GPS_data$lon), 
                                   lat = GPS_data$lat), proj4string = CRS("+proj=longlat"))

GPS_idx <- over(GPS_sp, Plist)
GPS_mu <- ice6g$trend[GPS_idx]
GPS_data$trend0 <- GPS_data$trend - GPS_mu
```

We also need to add some pseudo observations along the boudaries of the polygons to make smooth transition of the predictions. The values of these pseudo observations are set to be the ice6g values at the same locations with opposite signs.
```{r data2, include = TRUE, cache=TRUE}
## get the boundary of the polygons
boundlines <- as(zeroPoly, 'SpatialLines') 
obs_bounds <- spsample(boundlines, n = 200, type = "random") # note points more than specified
proj4string(obs_bounds) <- proj4string(Plist)
## Find the ice6g values
pobs_idx <- over(obs_bounds, Plist)
GIA_pobs <- ice6g$trend[pobs_idx]

nobsb <-nrow(obs_bounds@coords) 
obs_df <- data.frame(ID = rep("pseudo", nobsb), lon = obs_bounds@coords[,1], lat = obs_bounds@coords[,2],
                     trend = rep(0, nobsb), std = rep(0.05, nobsb), trend0 = -GIA_pobs)
obs_xyz <- do.call(cbind, Lll2xyz(lat = obs_bounds@coords[,2], lon = obs_bounds@coords[,1]))

GPS_all <- rbind(GPS_data, obs_df)
GPS_allxyz <- rbind(GPS_loc, obs_xyz)
```

## Prior setup for the parameters

Setup the priors for the parameters for the Gaussian process. Assume the prior distributions of both $\rho$ and $\sigma^2$ are log normal. The prior mean of the correlation length is set to be 500km based on expert prior opinion on the residual process and the mean for the variance is 20 which is about the range of the GIA values. The variances of both distributions are set to be large for a vague enough prior.
```{r prior}
## Priors mean and variance for the parameters: rho and sigma
mu_r <- 500/6371
v_r <- (1000/6371)^2
mu_s <- 20
v_s <- 40^2

## Transform the parameters for the SPDE_GMRF approximation
trho <- Tlognorm(mu_r, v_r)
tsigma <- Tlognorm(mu_s, v_s)
```


# INLA Inference on the subset

```{r inla, include = TRUE, cache = TRUE}
Mesh_GIA <- mesh_outPoly
## The SPDE model
lsigma0 <- tsigma[1]
theta1_s <- tsigma[2]
lrho0 <- trho[1]
theta2_s <- trho[2]
lkappa0 <- log(8)/2 - lrho0
ltau0 <- 0.5*log(1/(4*pi)) - lsigma0 - lkappa0
GIA_spde <- inla.spde2.matern(Mesh_GIA, B.tau = matrix(c(ltau0, -1, 1),1,3), B.kappa = matrix(c(lkappa0, 0, -1), 1,3),
                              theta.prior.mean = c(0,0), theta.prior.prec = c(sqrt(1/theta1_s), sqrt(1/theta2_s)))

## Link the process to observations and predictions
A_data <- inla.spde.make.A(mesh = Mesh_GIA, loc = GPS_allxyz)
A_pred <- inla.spde.make.A(mesh = Mesh_GIA, loc = rbind(GPS_allxyz, Mesh_GIA$loc))

## Create the estimation and prediction stack
st.est <- inla.stack(data = list(y=GPS_all$trend0), A = list(A_data),
                     effects = list(GIA = 1:GIA_spde$n.spde), tag = "est")
st.pred <- inla.stack(data = list(y=NA), A = list(A_pred),
                      effects = list(GIA=1:GIA_spde$n.spde), tag = "pred")
stGIA <- inla.stack(st.est, st.pred)

## Fix the GPS errors
hyper <- list(prec = list(fixed = TRUE, initial = 0))
formula = y ~ -1 +  f(GIA, model = GIA_spde)
prec_scale <- c(1/GPS_all$std^2, rep(1, nrow(A_pred)))
```

Then we run the INLA model. Note that this will take more than 10min and require memory larger than 32GB. We ran this on a server with 56 cores and 256GB memory.
```{r inla_run, include = TRUE, eval = FALSE}
## Run INLA
res_inla <- inla(formula, data = inla.stack.data(stGIA, spde = GIA_spde), family = "gaussian",
                 scale =prec_scale, control.family = list(hyper = hyper),
                 control.predictor=list(A=inla.stack.A(stGIA), compute =TRUE))

INLA_pred <- res_inla$summary.linear.predictor
```

```{r inla_load, include = FALSE, eval = TRUE}
#save(res_inla, file = "/./projects/GlobalMass/WP1-BHM/Experiment1b/GIA_RGL/res2.RData")

if(Sys.info()["sysname"] == "Windows"){
  load("Z:/WP1-BHM/Experiment1b/GIA_RGL/res2.RData")
}else if(grep("Ubuntu",Sys.info()["version"]) == 1){
 load("~/GMdata/GIA_RGL/res2.RData")
}
INLA_pred <- res_inla$summary.linear.predictor
```

# Analyse results

Now assemble the inla inference and prediction results.
```{r inla_res, include = TRUE, cache=TRUE}
## Extract and project predictions
pred_idx <- inla.stack.index(stGIA, tag = "pred")$data
GPS_idx <- pred_idx[1:nrow(GPS_all)]
GIA_idx <- pred_idx[-c(1:nrow(GPS_all))]

## GPS 
GPS_u <- INLA_pred$sd[GPS_idx]
GPS_pred <- data.frame(lon = GPS_all$lon, lat = GPS_all$lat, u = GPS_u)

## GIA
GIA_diff <- INLA_pred$mean[GIA_idx] 
GIA_m <- GIA_diff + GIA_prior[]
GIA_u <- INLA_pred$sd[GIA_idx]
proj <- inla.mesh.projector(Mesh_GIA, projection = "longlat", dims = c(360,180), xlim = c(0,360), ylim = c(-90, 90))
GIA_grid <- expand.grid(proj$x, proj$y)
GIA_pred <- data.frame(lon = GIA_grid[,1], lat = GIA_grid[,2],
                       diff = as.vector(inla.mesh.project(proj, as.vector(GIA_diff))),
                       mean = as.vector(inla.mesh.project(proj, as.vector(GIA_m))),
                       u = as.vector(inla.mesh.project(proj, as.vector(GIA_u))))

ress <- list(res_inla = res_inla, spde = GIA_spde, st = stGIA, 
            mesh = Mesh_GIA, GPS_pred = GPS_pred, GIA_pred = GIA_pred)
```


## Plot the posteriors of the hyper parameters

```{r hyper, include=TRUE}
res_inla <- ress$res_inla
GIA_spde <- ress$spde
pars_GIA <- inla.spde2.result(res_inla, "GIA", GIA_spde, do.transf=TRUE)
theta_mean <- pars_GIA$summary.theta$mean
theta_sd <- pars_GIA$summary.theta$sd

## Find the mode of rho and sigma^2
lrho_mode <- pars_GIA$summary.log.range.nominal$mode
lrho_mean <- pars_GIA$summary.log.range.nominal$mean
lrho_sd <- pars_GIA$summary.log.range.nominal$sd
rho_mode <- exp(lrho_mean - lrho_sd^2)

lsigma_mode <- pars_GIA$summary.log.variance.nominal$mode
lsigma_mean <- pars_GIA$summary.log.variance.nominal$mean
lsigma_sd <- pars_GIA$summary.log.variance.nominal$sd
sigma_mode <- exp(lsigma_mean - lsigma_sd^2)

plot(pars_GIA$marginals.range.nominal[[1]], type = "l",
     main = bquote(bold(rho("mode") == .(round(rho_mode, 4))))) # The posterior from inla output
plot(pars_GIA$marginals.variance.nominal[[1]], type = "l", 
     main = bquote(bold({sigma^2}("mode") == .(round(sigma_mode, 4))))) # The posterior from inla output

## The estimated correlation length is about 568km
rho_mode*6371
```


## Plot the predictions

```{r predict, include=TRUE}
GPS_pred <- ress$GPS_pred
GIA_pred <- ress$GIA_pred

GIA_pred$mean2 <- GIA_pred$mean
GIA_pred$mean2[is.na(GIA_pred$mean2)] <- 0

map_prior <- map_res(data = ice6g, xname = "x_center", yname = "y_center", fillvar = "trend", 
                      limits = c(-7, 22), title = "Prior GIA mean field")

## Plot the GIA predicted mean
map_GIA <- map_res(data = GIA_pred, xname = "lon", yname = "lat", fillvar = "mean", 
                   limits = c(-7, 22), title = "Predicted GIA")

map_GIA2 <- map_res(data = GIA_pred, xname = "lon", yname = "lat", fillvar = "mean2", 
                  limits = c(-7, 22), title = "Predicted GIA masked")

## Plot the GIA difference map
map_diff <- map_res(data = GIA_pred, xname = "lon", yname = "lat", fillvar = "diff", 
                   limits = c(-8, 8), title = "GIA difference: Updated - Prior")

## Plot the GIA difference map
map_sd <- map_res(data = GIA_pred, xname = "lon", yname = "lat", fillvar = "u", 
                    colpal = colorRamps::matlab.like(12),  title = "Predicted uncertainties")

 
## Display
print(map_prior)
print(map_GIA)
print(map_GIA2)
print(map_diff)
print(map_sd)
```