# Test INLA Mesh effects

***

We test the mesh effect on the marginal variances of Gaussian processes on $R^2$ and $S^2$ using the INLA package. The reason is that for a stationary Gaussian process, we would expect the marginal variances to be constant (at least within the boundary) however, we have observed non-constant variances: for irregular mesh, the differences are random and huge; while for regular mesh, the differences show certain pattern at the boundary or where the triangles change shapes.


Based on results (need references) of the finite element methods for SPDEs, we have the following statements

* The non-constant variance is caused by the shape change of the triangles.

* For regular mesh, when the triangles get small enough, the mesh effect should dis sapper.
    * How small is enough?
    * Relationship between maximum triangle edge length and the process correlation length.
    * Maximum edge < < correlation length? Or the other way round?

* For different high resolution meshes, the inferences on the parameters should be similar.

Therefore, in the following we tests the above statements first on $R^2$ then $S^2$. The first statement is obvious and we focus on the other two. Without loss of generality, we choose to test on a the square $[0,1] \times [0,1] \subset R^2$ and to reduce the boundary effects, the square is wrapped by a convex hull.

***

## 1. Marginal variances for a stationary process

### 1.1 Regular grid on $R^2$

We first test on $R^2$. The correlation length is fixed to be $\rho = 0.1$ and variance to be $\sigma^2 = 1$. The regular grid is generated by using equally spaced grid points and we increase the number of points in each dimension to have smaller triangles. To make sure the triangles are all the same within the boundary, we choose the maximum edge length to be $\sqrt{2}/l$, where $l$ is the number of segments in $x$ (or $y$) direction.


```{r warning=FALSE, message=FALSE}
library(INLA)
library(lattice)
library(fields)

## a wrapper for generating mesh, evaluating the variance and plot the result
plot.mvar <- function(range0, locres, max.edge, plot.mesh = TRUE, xylim = FALSE){
  sigma0 <- 1
  loc <- as.matrix(expand.grid(seq(0, 1, locres), seq(0, 1, locres)))
  mesh <- inla.mesh.2d(loc = loc, offset = c(0.1, 0.4), max.edge = max.edge )
  kappa0 <- sqrt(8)/range0
  tau0 <- 1/(sqrt(4*pi)*kappa0*sigma0)
  
  spde <- inla.spde2.matern(mesh, B.tau = cbind(log(tau0), -1, 1),
                            B.kappa = cbind(log(kappa0), 0, -1), 
                            theta.prior.mean = c(0,0), theta.prior.prec = c(0.1, 1))
  Q <- inla.spde.precision(spde, theta = c(0,0))
  msd <- sqrt(1/diag(Q))
  proj <- inla.mesh.projector(mesh, dims = c(100,100))
  
  if(xylim){
    image.plot(proj$x, proj$y, inla.mesh.project(proj, as.vector(msd)), 
             col = topo.colors(24), breaks = seq(0, 1.2, 0.05),
             xlim = c(0,1), ylim = c(0,1), xlab = "", ylab="",
             main = paste("edge length = ", locres))
  }else{
      image.plot(proj$x, proj$y, inla.mesh.project(proj, as.vector(msd)), 
                 col = topo.colors(24), breaks = seq(0, 1.2, 0.05),
                 xlab = "", ylab="",
                 main = paste("edge length = ", locres))
    }
  
  if(plot.mesh){
    plot(mesh, add = TRUE)}
}

### fix range and vary grid size
par(mfrow = c(2,3))
plot.mvar(range0=0.1, locres = 0.3, max.edge = c(0.45, 0.45))
plot.mvar(range0=0.1, locres = 0.2, max.edge = c(0.3, 0.3))
plot.mvar(range0=0.1, locres = 0.1, max.edge = c(0.15, 0.15))
plot.mvar(range0=0.1, locres = 0.08, max.edge = c(0.12, 0.12), plot.mesh = FALSE)
plot.mvar(range0=0.1, locres = 0.04, max.edge = c(0.06, 0.06), plot.mesh = FALSE)
plot.mvar(range0=0.1, locres = 0.02, max.edge = c(0.03, 0.03), plot.mesh = FALSE)
```


#### Remarks

The true standard error should be equal to 1 and the closest in the above plots is the those with edge length around 0.1 which is also the correlation length of the process. The standard error is under approximated using either a low or high resolution grid. 

**_We may conclude that the best resolution of the mesh would be the one with triangle length approximately equal to the process correlation length._**


### 1.2 Semi-regular grid on $S^2$
Next we do the same text on $S^2$. Note that on $S^2$ it is not possible to generate a mesh that is entirely regular. We used the INLA function to generate a semi-regular mesh based on evenly segmenting the edges of an icosahedron.


```{r warning=FALSE, message=FALSE}

## a wrapper for generating mesh on the sphere, evaluating the variance and plot the result
plot.mvarS2 <- function(range0, nseg){
  sigma0 <- 1
  mesh <- inla.mesh.create(globe = nseg)
  kappa0 <- sqrt(8)/range0
  tau0 <- 1/(sqrt(4*pi)*kappa0*sigma0)
  
  spde <- inla.spde2.matern(mesh, B.tau = cbind(log(tau0), -1, 1),
                            B.kappa = cbind(log(kappa0), 0, -1), 
                            theta.prior.mean = c(0,0), theta.prior.prec = c(0.1, 1))
  Q <- inla.spde.precision(spde, theta = c(0,0))
  msd <- sqrt(1/diag(Q))
  proj <- inla.mesh.projector(mesh, dims = c(360,181))

  image.plot(proj$x, proj$y, inla.mesh.project(proj, as.vector(msd)), 
             col = topo.colors(40), breaks = seq(0,1,0.025),
             xlab = "", ylab="",
             main = paste("edge length = ", 1/nseg))
}

### fix range and vary grid size
par(mfrow = c(2,3))
plot.mvarS2(range0=0.1, nseg = 2)
plot.mvarS2(range0=0.1, nseg = 5)
plot.mvarS2(range0=0.1, nseg = 10)
plot.mvarS2(range0=0.1, nseg = 20)
plot.mvarS2(range0=0.1, nseg = 50)
plot.mvarS2(range0=0.1, nseg = 100)
```

#### Remarks

The mesh effect due to the shape change of triangles can not be eliminated but the variation becomes negligible when the edge length of the triangles are similar to the correlation length and the variance approximation is also more accurate.

***

## 2. Inference on the parameters

Now we test the mesh effect on parameter inference.

### 2.1 On $R^2$

Here we compare posterior marginal distributions of the hyper parameters, $\rho$ and $\sigma^2$ from four different meshes using simulated data. The four meshes include three regular meshes with maximum edge length larger, approximately equal and smaller than the correlation length and an irregular mesh with maximum edge length approximately equal to the correlation length.

In the following experiment, we assume the true process has $\rho = 0.1, \sigma^2 = 1$ and we study the region $[0,1] \times [0,1] \subset R^2$.

```{r warning=FALSE, message=FALSE}
## Simulate 500 data points from the true process
data_loc <- matrix(runif(500*2), 500, 2)
data_cov <- Matern()

```
